---
title: "The Very Basics of R"
author: "mf"
date: "November 3, 2017"
output: 
    beamer_presentation:
        theme: "Madrid"
        
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
## Today's plan  

* 1) Getting around in R studio

* 2) Wow! 

* 3) Boring syntax stuff 

* 4) Regressions


## Why R? (R vs. STATA)  

* R does a bunch of things better than STATA  
    - Web stuff  
    - Manipulating weird data formats  
    - Graphics 
    - Machine learning  
    - Big Data
    
* R is much more commonly used by data scientists
    - (also Python) 
    
* R is free 

* However: 
    - R is way harder to learn
    - R is much less practical for econometrics 


## Getting started with R Studio
* R studio is not R
    - Open Rx64 or Ri368 program (just to see it)
    - We'll never do this again. Just use R Studio. 
    
* Open R studio 
    - 4 Panes: Source (code) / Console (output) / 2 more for a bunch of things
    - Make a new project (File/NewProject) - set the directory to wherever u want
    - Open a new R script
    - Type "5" and hit control+Enter on the keyboard   
```{r five, echo = TRUE}
# When I write "5" in the code and press Ctrl+Enter
5
# is the output in the console
```

## How R differs from Stata in usage 
* R functions with "Work Spaces" associated with a project file .Rprj
* R holds many objects/datasets at once in the "environment"
```{r dsets, echo = TRUE}
data(cars); data(state)
```
* You usually run R programs line-by-line 
* Everything depends on packages
    - kinda like ado files, but not exactly 
    - packages contain a bunch of functions  
(not like ado file, which only contains one command)
    - tons of flexibility 
    - all made by uncoordinated private users: very little consistency in syntax
    - often conflicts between packages
* You have to load packages every time into your "workspace"
   
    
## Packages 
```{r packages, echo = TRUE, eval=FALSE}
# You install packages with the command install.packages(): 
install.packages("dplyr")
# Then the packages exist in your local library (see panel)  

# But they are not loaded until you run the library() command: 
library(dplyr)
library(plyr)

# There are lots of competing/conflicting packages: 
library(xlsx)
library(xlsx2)
```


## The very basics of R syntax 
```{r syntax, echo = TRUE}
# 1) Comments start with a Hashtag. No multiline comments. 
# 2) Operations get resolved (i.e. you will see 16, not "8+8") 
8+8 
sqrt(244)
# 3) Assignments are done with " <- " or " = " 
a <- 4  
b = 20  
a + b

```

## The very basics of R syntax (2)
```{r syntax2, echo = TRUE}
# 4) Commands figure out line breaks automatically   
20 +  
sqrt(
        4)

# 5) But you can also use ";"  
c <- 1; d <-2; e <-3 ; 
c + d + e 
```

## The very basics of R syntax (3)
```{r syntax3, echo = TRUE}
# 6) vectors are defined with the c() function 
c(1,2,3,4)

# 7) R thinks of most objects as matrices (or arrays)  
c1 <- c(1,2,3,4)
c2 <- c(2,6,5,4)
cbind(c1,c2)
```

## The very basics of R syntax (4)
```{r syntax4, echo = TRUE}
# 8) You can mix data types (sortof)... 
v <- c(1,3*2,3,"J. Bieber")  
v  

# 9) Operators are mostly like stata 
# ==    is logical equal 
# !     means NOT
# &     means AND
# |     means OR
# (there are more) 

# 10) We'll get back to syntax later. 
```


