---
title: "Syntax Stuff (from coursera)"
author: "mf"
date: "November 3, 2017"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
## Source of this material (and further reading)

Coursera class "R programming"
from Johns Hopkins University, 
by Roger Peng, Jeff Leek, and Brian Caffo. 
https://www.coursera.org/learn/r-programming/ 


All their materials are on github at
https://github.com/DataScienceSpecialization/courses/tree/master/02_RProgramming



## Objects

R has five basic or “atomic” classes of objects:

-   character

-   numeric (real numbers)

-   integer

-   complex

-   logical (True/False)

The most basic object is a vector

-   A vector can only contain objects of the same class

-   BUT: The one exception is a *list*, which is represented as a vector
    but can contain objects of different classes (indeed, that’s usually
    why we use them)
---    
    
## Attributes 
R objects can have attributes
-   names, dimnames

-   dimensions (e.g. matrices, arrays)

-   class

-   length

-   other user-defined attributes/metadata

Attributes of an object can be accessed using the ```attributes()```
function.

## Creating Vectors

The `c()` function can be used to create vectors of objects.

```r
> x <- c(0.5, 0.6)       ## numeric
> x <- c(TRUE, FALSE)    ## logical
> x <- c(T, F)           ## logical
> x <- c("a", "b", "c")  ## character
> x <- 9:29              ## integer
> x <- c(1+0i, 2+4i)     ## complex
```
If you mix objects, R will force them to all be the same class (usually strings)

## Matrices

Matrices are vectors with a _dimension_ attribute. The dimension attribute is itself an integer vector of length 2 (nrow, ncol)

```{r , echo=TRUE}
m <- matrix(nrow = 2, ncol = 3) 
m
dim(m)
attributes(m)
```

## cbind-ing and rbind-ing

Matrices can be created by _column-binding_ or _row-binding_ with `cbind()` and `rbind()`.

```r
> x <- 1:3
> y <- 10:12
> cbind(x, y)
     x  y 
[1,] 1 10 
[2,] 2 11 
[3,] 3 12
> rbind(x, y) 
  [,1] [,2] [,3]
x    1    2    3
y   10   11   12
```

## Lists

Lists are a special type of vector that can contain elements of different classes. Lists are a very important data type in R and you should get to know them well.

```r
> x <- list(1, "a", TRUE, 1 + 4i) 
> x
[[1]]
[1] 1

[[2]] 
[1] "a"

[[3]]
[1] TRUE

[[4]]
[1] 1+4i
```

## Missing Values

Missing values are denoted by `NA` or `NaN` for undefined mathematical operations. 

- `is.na()` is used to test objects if they are `NA`

- `is.nan()` is used to test for `NaN` (Not A Number (like divided by 0 etc.))

- `NA` values have a class also, so there are integer `NA`, character `NA`, etc.

- A `NaN` value is also `NA` but the converse is not true

---

## Missing Values

```r
> x <- c(1, 2, NA, 10, 3)
> is.na(x)
[1] FALSE FALSE  TRUE FALSE FALSE
> is.nan(x)
[1] FALSE FALSE FALSE FALSE FALSE
```

## Data Frames

Data frames are used to store tabular data

- They are represented as a special type of list where every element of the list has to have the same length

- Each element of the list can be thought of as a column and the length of each element of the list is the number of rows

- Unlike matrices, data frames can store different classes of objects in each column (just like lists); matrices must have every element be the same class

- Data frames also have a special attribute called `row.names`

- Data frames are usually created by calling `read.table()` or `read.csv()`

- Can be converted to a matrix by calling `data.matrix()`

---

## Data Frames

```r
> x <- data.frame(foo = 1:4, bar = c(T, T, F, F)) 
> x
  foo   bar
1   1  TRUE
2   2  TRUE
3   3 FALSE
4   4 FALSE
> nrow(x)
[1] 4
> ncol(x)
[1] 2
```
---

## Reading Data

There are a few principal functions reading data into R.   
- `read.table`, `read.csv`, for reading tabular data   
- `readLines`, for reading lines of a text file  
- `source`, for reading in R code files (`inverse` of `dump`)   
- `dget`, for reading in R code files (`inverse` of `dput`)  
- `load`, for reading in saved workspaces  
- `unserialize`, for reading single R objects in binary form 

NOTE: help on any function can be looked up with ?_name_of_function_

--- 

## Writing Data

There are analogous functions for writing data to files  
- write.table  
- writeLines  
- dump  
- dput  
- save  
- serialize  

---

## Reading Data Files with read.table

The `read.table` function is one of the most commonly used functions for reading data. It has a few important arguments:  
- `file`, the name of a file, or a connection  
- `header`, logical indicating if the file has a header line  
- `sep`, a string indicating how the columns are separated  
- `colClasses`, a character vector indicating the class of each column in the dataset   
- `nrows`, the number of rows in the dataset  
- `comment.char`, a character string indicating the comment character  
- `skip`, the number of lines to skip from the beginning  
- `stringsAsFactors`, should character variables be coded as factors?  


---
## Interfaces to the Outside World

Data are read in using _connection_ interfaces. Connections can be made to files (most common) or to other more exotic things.

- `file`, opens a connection to a file  
- `gzfile`, opens a connection to a file compressed with gzip  
- `url`, opens a connection to a webpage  

In general, connections are powerful tools that let you navigate files or other external objects. In practice, we often don’t need to deal with the connection interface directly.
```r
con <- file("foo.txt", "r")
data <- read.csv(con)
close(con)
```
is the same as

```r
data <- read.csv("foo.txt")
```

## Reading Lines of a Text File

`readLines` can be useful for reading in lines of webpages

```r
## This might take time
con <- url("http://www.ifpri.org", "r")
x <- readLines(con)
head(x,10)
close(con)
```

## Subsetting

There are a number of operators that can be used to extract subsets of R objects.  

- `[` always returns an object of the same class as the original; can be used to select more than one element  

- `[[` is used to extract elements of a list or a data frame; it can only be used to extract a single element and the class of the returned object will not necessarily be a list or data frame  

- `$` is used to extract elements of a list or data frame by name; semantics are similar to that of `[[`.  


## Subsetting 2 

```r
> x <- c("a", "b", "c", "c", "d", "a")
Try all of the following: 
> x[1]
> x[2]
> x[1:4]
> x[x > "a"]

and some more: 
> u <- x > "a"
> u
> x[u]
```

## Subsetting a Matrix 
(This is starting to be useful for data manipulation)   
Matrices can be subsetted in the usual way with (_i,j_) type indices.

```r
> x <- matrix(1:6, 2, 3)
> x[1, 2]
> x[2, 1]
Missing index means the whole row/column:  
> x[1, ]
> x[, 2]
```

## Subsetting Lists
```r
> x <- list(foo = 1:4, bar = 0.6)
> x[1]
$foo
[1] 1 2 3 4

> x[[1]]
[1] 1 2 3 4

> x$bar
[1] 0.6
> x[["bar"]]
[1] 0.6
> x["bar"]
$bar
[1] 0.6
```

## Removing NA Values

A common task is to remove missing values (`NA`s).

```r
> x <- c(1, 2, NA, 4, NA, 5)
> bad <- is.na(x)
> x[!bad]
[1] 1 2 4 5
```

```r
> x <- c(1, 2, NA, 4, NA, 5)
> y <- c("a", "b", NA, "d", NA, "f")
> good <- complete.cases(x, y)
> good
[1]  TRUE  TRUE FALSE  TRUE FALSE  TRUE
> x[good]
[1] 1 2 4 5
> y[good]
[1] "a" "b" "d" "f"
```


## Removing NA Values

```r
> airquality[1:6, ]
  Ozone Solar.R Wind Temp Month Day
1    41     190  7.4   67     5   1
2    36     118  8.0   72     5   2
3    12     149 12.6   74     5   3
4    18     313 11.5   62     5   4
5    NA     NA  14.3   56     5   5
6    28     NA  14.9   66     5   6
> good <- complete.cases(airquality)
> airquality[good, ][1:6, ]
  Ozone Solar.R Wind Temp Month Day
1    41     190  7.4   67     5   1
2    36     118  8.0   72     5   2
3    12     149 12.6   74     5   3
4    18     313 11.5   62     5   4
7    23     299  8.6   65     5   7
```


## How R does looping

`apply` is used to a evaluate a function (often an anonymous one) over the margins of an array.  

- It is most often used to apply a function to the rows or columns of a matrix  

- It can be used with general arrays, e.g. taking the average of an array of matrices   

- It is not really faster than writing a loop, but it works in one line!  

apply(_data_, _dimension_, _function_)

For sums and means of matrix dimensions, we have some shortcuts.
- `rowSums` = `apply(x, 1, sum)`  
- `rowMeans` = `apply(x, 1, mean)`  
- `colSums` = `apply(x, 2, sum)`  
- `colMeans` = `apply(x, 2, mean)`  

## Other "apply-type" functions: 

- `lapply`: Loop over a list and evaluate a function on each element  

- `sapply`: Same as `lapply` but try to simplify the result  

- `apply`: Apply a function over the margins of an array  
 
- `tapply`: Apply a function over subsets of a vector  

- `mapply`: Multivariate version of `lapply`  


## For, while, etc. 
Those also exist, but the `apply` functions are what is new for us stata people.  

```{r echo=TRUE}  
for (year in c(2010,2011,2012,2013,2014,2015)){
    print(paste("The year is", year))
}
```

## Further training 

These slides were cherry-picked material from a 4-week course.  

Take it if you want more practice:  
Coursera class "R programming" from Johns Hopkins University,   
by Roger Peng, Jeff Leek, and Brian Caffo.  
https://www.coursera.org/learn/r-programming/  


Or just read their slides 
https://github.com/DataScienceSpecialization/courses/tree/master/02_RProgramming



